Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2022-11-30 paper/tidy-missing-data-paper.Rmd
small copyedits

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, "\\pkg{tidyverse}", data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
    includes:
      header-includes: 
        - \usepackage{amsmath}
        - \usepackage[english]{babel}
        - \usepackage{float}
        - \usepackage{booktabs}
bibliography: "tidy-missing-data-zotero.bib"
cls: jss.cls
---

@@

While there have been many software tools for exploring missing data, they do not work together. The graphics literature provides several solutions for exploring missings visually by incorporating them into the plot in some way. For example, imputing values to be 10% below the minimum to include all observations into a scatter plot [@Cook2007], or treating missing values as an equal category of data [@Unwin1996]. The ideas from the graphics literature need to be translated into tidy data tools to integrate missing data handling in a data analysis pipeline.

@@

$$
b_{ij} =\begin{cases}
  1, & \text{if } y_{ij} \text{ is missing} \\
  0, & \text{if } y_{ij} \text{ is observed}
\end{cases}
$$

@@

```{r make-school-scores-data, echo = FALSE}

school_scores <- data.frame(score = c(rnorm(20, 70, 10),
                                 rnorm(20, 85, 7.5),
                                 rnorm(20, 95, 5),
                                 rnorm(20, 115, 10)),
                       year = rep(c("1st", "2nd", "3rd", "4th"), each = 20)) %>%
  mutate(score = as.numeric(score),
         year = as.factor(year)) %>%
  as_tibble()

>>>>

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar} \citep{pkg:naniar}, available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/package=naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, "\\pkg{tidyverse}", data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
    includes:
      header-includes: 
        - \usepackage{amsmath}
        - \usepackage[english]{babel}
        - \usepackage{float}
        - \usepackage{booktabs}
bibliography: "tidy-missing-data-zotero.bib"
cls: jss.cls
---

@@

While there have been many software tools for exploring missing data, they do not work together. The graphics literature provides several solutions for exploring missings visually by incorporating them into the plot in some way. For example, imputing values to be 10% below the minimum to include all observations into a scatter plot [@Cook2007], or treating missing values as an equal category of data [@Unwin1996]. The ideas from the graphics literature need to be translated into tidy data tools to integrate missing data handling in a data analysis pipeline. We have implemented these tools for exploring missing data into the \proglang{R} package \pkg{naniar} \citep{pkg:naniar}, available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/package=naniar}.

@@

$$
b_{ij} = \begin{cases}
  1, \text{if } y_{ij} \text{ is missing} \\
  0, \text{if } y_{ij} \text{ is observed}
\end{cases}
$$

@@

```{r make-school-scores-data, echo = FALSE}
set.seed(2022-11-29)
school_scores <- data.frame(score = c(rnorm(20, 70, 10),
                                 rnorm(20, 85, 7.5),
                                 rnorm(20, 95, 5),
                                 rnorm(20, 115, 10)),
                       year = rep(c("1st", "2nd", "3rd", "4th"), each = 20)) %>%
  mutate(score = as.numeric(score),
         year = as.factor(year)) %>%
  as_tibble()
